---
title: "Sample size simulation"
author: "Han Yu"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(causalTree)
library(purrr)
library(grf)
library(dplyr)
library(tidyverse)
library(parallel)
library(iterators)
library(doParallel)
library(foreach)
library(ggplot2)
library(plotly)
```

## Define the functions

### Generate data

Suppose we have a data set with $n$ independently and identically distributed (iid) units with $i = 1, ···, n$. Each unit has an observed feature vector $X_i \in [0,1]^d$, with $d$ binary covariates and a response (the outcome of interest) $Y_i \in \{0,1\}$, where $x_1,···,x_k$ are effect modifiers (covariates contribute to the treatment effect), and $x_{k+1},···,x_d$ are other risk factors/noise covariates (covariates act as a distractor to the identification of variables that are relevant for the treatment effect).

Treatments are considered binary and formalized with an indicator variable $A_i \in \{0,1\}$ for each unit $i$. Treatments are randomly assigned to the whole population, and random errors are defined as $\varepsilon \sim N (0, \sigma_0^2)$.

To do this, we assume an additive model and follow a bi-partite data generation process, in which two equations are used (one for treated cases and another for control cases).

For the treated cases, we use $$\begin{equation*} P(Y_{i}(1)) = \beta_0 + \sum_{j=1}^{d} \beta_{j}x_{ij} + A_{i}\ast (\gamma_0+\prod_{j=1}^{k} \gamma x_{ij}) + \varepsilon = \beta_0 + \sum_{j=1}^{d} \beta_{j}x_{ij} + \gamma_0+\prod_{j=1}^{k} \gamma x_{ij} + \varepsilon \end{equation*}$$ with treatment indicator variable $A_i$ set to 1.

$$Y_i(1) \mid X \sim Bernoulli(logit^{-1}(\beta_0 + \sum_{j=1}^{d} \beta_{j}x_{ij} + \gamma_0+\prod_{j=1}^{k} \gamma x_{ij} + \varepsilon))$$

For the control cases, we use$$\begin{equation*} P(Y_{i}(0)) = \beta_0 + \sum_{j=1}^{d} \beta_{j}x_{ij} + A_{i}\ast (\gamma_0+\prod_{j=1}^{k} \gamma x_{ij}) + \varepsilon = \beta_0 +  \sum_{j=1}^{d} \beta_{j}x_{ij}+\varepsilon \end{equation*}$$with treatment indicator variable $A_i$ set to 0.

$$Y_i(0) \mid X \sim Bernoulli(logit^{-1}(\beta_0 + \sum_{j=1}^{d} \beta_{j}x_{ij} + \varepsilon))$$

For a unit-level causal effect, the heterogeneous causal effect on unit $i$ is defined as $\tau(x_i) = E[Y_i(1)-Y_i(0) \mid X_i = x]$, the difference between treated $Y_i(1)$ and untreated $Y_i(0)$ outcome. The ground truth of $\tau$ is $$\tau(x_i) = E[Y_i(1)-Y_i(0) \mid X_i = x] = E[Y_i(1)\mid X_i = x] - E[Y_i(0) \mid X_i=x] = P(Y_i(1)) - P(Y_i(0)) = \gamma_0 + \prod_{j=1}^{k}\gamma x_{ij}$$

```{r}
generate_data <- function(seed = 42, n = 1000, k = 1, d = 2, x_prob = NULL, beta = NULL, gamma = NULL, noise_sd = 0.1) {
  
  # Input parameter
  # n: sample size
  # k: number of effect modifiers
  # d: number of total covariates (the first k are effect modifiers)
  # x_prob: the probability that each covariate equal to 1 (parameters for the Bernoulli distribution)
  # beta: Coefficients controlling the effect of covariates on the baseline risk (including intercept)
  # gamma: Coefficients controlling the effect of effect modifiers on the treatment effect (including main effect)
  # noise_sd: Standard deviation of the Gaussian noise term
  # seed: Random seed for reproducibility
  
  # Set random seed for reproducibility
  set.seed(seed)
  
  ############## Input check and eror print ############## 
  
  # set default value for x_prob and error check
  if (is.null(x_prob)) x_prob <- rep(0.5, d)
  
  if (!is.numeric(x_prob)) stop("Error: The 'x_prob' parameter must be a numeric vector.")
  
  if (length(x_prob) != d) stop("Error: The length of 'x_prob' must equal the total number of covariates.")
  
  # set default value for beta and error check
  if (is.null(beta)) beta <- rep(1, d+1)
  
  if (!is.numeric(beta)) stop("Error: The 'beta' parameter must be a numeric vector.")
  
  if (length(beta) != d+1) stop("Error: The length of 'beta' must equal the total number of covariates + 1 (main baseline risk).")
  
  # set default value for gamma and error check
  # if (is.null(gamma)) gamma <- rep(1, k+1)
  # 
  # if (!is.numeric(gamma)) stop("Error: The 'gamma' parameter must be a numeric vector.")
  # 
  # if (length(gamma) != k+1) stop("Error: The length of 'gamma' must equal the number of effect modifiers + 1 (main effect).")
  
  # error check
  if (d < k) stop("Error: The total number of the covariates should be larger than the number of effect modifiers.")
  
  if(k > 5) stop("Error: The maximum number of the effect modifier is 5.")
  
  #########################################################
  
  # Generate n observations of d binary covariates
  X <- sapply(x_prob, function(p) rbinom(n, size = 1, prob = p))
  colnames(X) <- paste0("X", 1:d)
  
  # Generate phenotype      
  phenotype <- as.integer(apply(X[, 1:k, drop = FALSE], 1, prod) == 1) 
  
  # Random treatment assignment A (Bernoulli(0.5))
  A <- rbinom(n, 1, 0.5)
  
  # Random noise (epsilon), normally distributed
  epsilon <- rnorm(n, mean = 0, sd = noise_sd)
  
  # Compute potential outcomes
  Y1_linear <- beta[1] + X %*% beta[2:(d+1)] + gamma[1] + apply(X[, 1:k, drop = FALSE], 1, prod) * gamma[2] + epsilon
  # Y1_linear <- beta[1] + X %*% beta[2:(d+1)] + gamma[1] + X[, 1:k, drop = FALSE] %*% gamma[2:(k+1)] + epsilon
  
  Y0_linear <- if (d-k > 0) beta[1] + X %*% beta[2:(d+1)] + epsilon else epsilon
  
  p1 <- plogis(Y1_linear)
  p0 <- plogis(Y0_linear)
  
  # Observed outcome Y depends on treatment assignment A
  Y1 <- rbinom(n, 1, p1)
  Y0 <- rbinom(n, 1, p0)
  #Y <- A * Y1 + (1-A) *Y0
  Y <- ifelse(A==1, Y1, Y0)
  
  # Return the dataset as a data frame
  data <- data.frame(
    ID = 1:n,
    A = A,
    Y = Y,
    #Y1 = Y1,
    #Y0 = Y0,
    CATE = p1-p0
  )
  
  data <- cbind(data, as.data.frame(X), as.data.frame(phenotype))
  
  return(data)
}
```

```{r}
sim_data <- generate_data(seed = 42, n = 1000, k = 1, d = 2, x_prob = c(0.5,0.7), beta = c(0, 0.1, 0.1), gamma = c(0, 1.6), noise_sd = 0.1)
sim_data2 <- generate_data(seed = 42, n = 1000, k = 2, d = 3, x_prob = c(0.5,0.5,0.7), beta = c(0, 0.1, 0.1, 0.1), gamma = c(0, 1.6), noise_sd = 0.1)
```

### Percentage of trees have at least one pure X (effect modifier) leaf among the forest

```{r}
check_trees_purity <- function(forest, data, num_trees, mode = c("x","phenotype")) {
  mode <- match.arg(mode)
  
  success <- 0
  unsuccess <- 0
  partial_success <- 0
  
  for(i in seq_len(num_trees)) {
    
    leaf_list <- forest$`_leaf_samples`[[i]]
    leaf_nodes <- lapply(leaf_list, function(x) length(x) > 0)
    leaf_node_indices <- leaf_list[unlist(leaf_nodes)]

    if (mode == "x"){
      
      x1_props <- sapply(leaf_node_indices, function(leaf) mean(data$X1[leaf + 1]))
  
      if (all(x1_props %in% c(0,1))){
        success <- success + 1
        #next
      } else if (any(x1_props == 1) && any(x1_props > 0 & x1_props < 1)) {
        partial_success <- partial_success + 1
      } else{
        unsuccess <- unsuccess + 1
      }
      
    } else if(mode == "phenotype"){
      
      # number of samples with phenotype == 1
      all_leaf_samples <- unlist(leaf_node_indices) + 1
      n_current_tree_all_pheno <- sum(data[all_leaf_samples, ]$phenotype)
      
      # if no phenotype==1 sample, skip this tree
      if (n_current_tree_all_pheno == 0) {
        unsuccess <- unsuccess + 1
        next
      }
      
      is_full_phenotype <- vapply(leaf_node_indices, function(leaf) {
        sample_indices <- leaf + 1
        all(data[sample_indices,]$phenotype == 1)
      }, logical(1))
      
      phenotype_covered <- sum(vapply(leaf_node_indices[is_full_phenotype], length, integer(1)))
      prop_phenotype <- phenotype_covered / n_current_tree_all_pheno
  
      if (prop_phenotype == 1) {
        success <- success + 1
      } else if (prop_phenotype > 0) {
        partial_success <- partial_success + 1
      } else {
        unsuccess <- unsuccess + 1
      }
      
    }
    
  }
  
  return(list(
    success_ratio = success / num_trees,
    partial_success_ratio = partial_success / num_trees,
    unsuccess_ratio = unsuccess / num_trees
    ))
}
```

```{r}
data <- generate_data(seed = 123, n = 1000, 
                      k = 1, d = 2, 
                      x_prob = c(0.5,0.7),
                      beta = c(0, 0.1, 0.1),
                      gamma = c(0, 1.6))
forest <- causal_forest(data %>% select(starts_with("X")), 
                                         data$Y, 
                                         data$A,
                                         num.trees = 1000,
                                         sample.fraction = 0.5,
                                         honesty = TRUE,
                                         honesty.fraction = 0.5,
                                         seed = 123)
```

### Heterogeneity captured by X (effect modifier)

The percentage of heterogeneity captured among the forest is defined as following:

$$ \begin{equation*}
\frac{\frac{1}{n_1}\sum\limits_{i \in S_1}CATE(\prod_{j=1}^{k} X_{ij}=1;\underset{\sim}{X_{-ij}})
\;-\;
\frac{1}{n_0}\sum\limits_{i \in S_0}CATE(\prod_{j=1}^{k} X_{ij}=0;\underset{\sim}{X_{-ij}})}{ground \ truth \ of \ CATE}
\;\times\;
100\%
\end{equation*} $$

, where $S_1$ is defined as the subset of $\prod_{j=1}^{k}X_{ij}=1$, $S_0$ is defined as the subset of $\prod_{j=1}^{k}X_{ij}=0$, and ground truth of CATE can be obtained from the data simulation process.

### Calculate the above two index

```{r}
# reference of causal_forest function:
# https://grf-labs.github.io/grf/REFERENCE.html

forest_built <- function(data, 
                         num_trees, 
                         k, 
                         sample.fraction = 0.5, 
                         honesty = TRUE, 
                         honesty.fraction = 0.5){
  
  # Train Causal Forest
  if(honesty == FALSE){
    causal_forest_model <- causal_forest(data %>% select(starts_with("X")), 
                                         data$Y, 
                                         data$A,
                                         num.trees = num_trees,
                                         sample.fraction = sample.fraction,
                                         honesty = honesty,
                                         seed = 123)
  }else{
    causal_forest_model <- causal_forest(data %>% select(starts_with("X")), 
                                         data$Y, 
                                         data$A,
                                         num.trees = num_trees,
                                         sample.fraction = sample.fraction,
                                         honesty = honesty,
                                         honesty.fraction = honesty.fraction,
                                         seed = 123)
  }
  
  # predicted cate
  tau.hat.est <- causal_forest_model$predictions
  
  # heterogeneity captured
  pheno_data_index <- data$phenotype == 1
    
  truth_cate <- mean(data$CATE[pheno_data_index]) - mean(data$CATE[!pheno_data_index])
    
  cate_phenotype_1 <- mean(tau.hat.est[pheno_data_index])
  cate_phenotype_0 <- mean(tau.hat.est[!pheno_data_index])
  heterogeneity_captured <- ((cate_phenotype_1 - cate_phenotype_0) / truth_cate)
  
  # only one effect modifier
  if(k==1){
    tmp_result <- check_trees_purity(causal_forest_model, data, num_trees, mode="x")
  }else{ # more than 1 effect modifiers
    tmp_result <- check_trees_purity(causal_forest_model, data, num_trees, mode="phenotype")
  }
  
  success_ratio <- tmp_result[[1]]
  partial_success_ratio <- tmp_result[[2]]
  unsuccess_ratio <- tmp_result[[3]]
    
  return(list(heterogeneity_captured = heterogeneity_captured,
              success_ratio = success_ratio,
              partial_success_ratio = partial_success_ratio,
              unsuccess_ratio = unsuccess_ratio))
  
}
```

### Calculate the overall sample size

```{r}
run_simulation <- function(sim_no = 500, 
                           sample_param = NULL, 
                           param = NULL,
                           k = 1, 
                           d = 2, 
                           x_prob, 
                           beta, 
                           gamma, 
                           num_trees = 2000){
  
  is_tune_mode <- !is.null(param)
  is_sample_mode <- !is_tune_mode
  
  if (is_tune_mode && length(sample_param) != 1) {
  stop("In tuning mode, sample size must be a single fixed number.")
}
  
  if (is_sample_mode) {
    param_grid <- data.frame(sample_size = sample_param)
  } else {
    param_grid <- expand.grid(param)
  }
  
  n_param_grid <- nrow(param_grid)
  
  heterogeneity <- numeric(n_param_grid)
  success <- numeric(n_param_grid)
  partial_success <- numeric(n_param_grid)
  unsuccess <- numeric(n_param_grid)

  # Detect cores and set up parallel backend
  num_cores <- detectCores() - 1
  cl <- makeCluster(num_cores)
  registerDoParallel(cl)

  for(i in seq_len(n_param_grid)){
    start_time <- Sys.time()
    set.seed(1)
    seeds <- sample(1:(sim_no*2), sim_no, replace = FALSE)

    # Parallel loop over simulations
    results <- foreach(j = 1:sim_no,
                       .combine = rbind,
                       .packages = c("grf", "dplyr"),
                       .export = c("generate_data", 
                                   "forest_built",
                                   "check_trees_purity")) %dopar% {

      sim_data <- generate_data(
        seed = seeds[j], 
        n = if (is_sample_mode) param_grid$sample_size[i] else sample_param, 
        k = k, 
        d = d, 
        x_prob = x_prob, 
        beta = beta, 
        gamma = gamma
        )

      # Build forest
      result <- forest_built(
        data = sim_data, 
        num_trees = num_trees, 
        k = k,
        sample.fraction = if (!is_sample_mode) param_grid[i, 1] else 0.5,
        honesty.fraction = if (!is_sample_mode) param_grid[i, 2] else 0.5
        )

      # Combine results as a row
      c(result$heterogeneity_captured, 
        result$success_ratio, 
        result$partial_success_ratio, 
        result$unsuccess_ratio)
      }
      

    # Compute the average/median from the parallel results
    heterogeneity[i] <- median(results[, 1])
    success[i] <- median(results[, 2])
    partial_success[i] <- median(results[, 3])
    unsuccess[i] <- median(results[,4])
    
    end_time <- Sys.time()
    if (is_sample_mode) {
      cat("Sample size", param_grid$sample_size[i], ": running time", end_time - start_time, "\n")
    } else {
      cat("Tuning param", paste(names(param_grid), unlist(param_grid[i, ]), sep = "=", collapse = ", "), ": running time", end_time - start_time, "\n")
    }
  }
  
  stopCluster(cl)
  
  return_result <- data.frame(
    heterogeneity_captured = heterogeneity,
    success_ratio = success,
    partial_success_ratio = partial_success,
    unsuccess_ratio = unsuccess
    )
  
  return_result <- cbind(param_grid, return_result)
  return(return_result)
}
```

## Results

### Conduct the simulation study (tab 1)

#### One effect modifier X1 + One noise covariate X2

```{r}
sim_result1 <- run_simulation(sim_no = 500,
                              sample_param = c(100,200,400,600,800,1000,1200),
                              #sample_param = c(100),
                              k = 1, d = 2,
                              x_prob = c(0.5,0.7),
                              beta = c(0, 0.1, 0.1),
                              gamma = c(0, 1.6),
                              num_trees = 1000)
```

#### Two effect modifiers X1, X2 + One noise covariate X3

```{r}
sim_result2 <- run_simulation(sim_no = 500,
                              sample_param = c(100,200,400,600,800,1000,1200),
                              #sample_param = c(100,200,400,600),
                              k = 2, d = 3,
                              x_prob = c(0.5,0.5,0.7),
                              beta = c(0, 0.1, 0.1,0.1),
                              gamma = c(0, 1.6),
                              num_trees = 1000)
```

#### Two effect modifiers X1, X2 + Two noise covariates X3, X4

```{r}
sim_result3 <- run_simulation(sim_no = 500,
                              #sample_param = c(100,200,400,600,800,1000,1200),
                              sample_param = c(100,200,400,600),
                              k = 2, d = 4,
                              x_prob = c(0.5,0.5,0.5,0.7),
                              beta = c(0, 0.1, 0.1, 0.1, 0.1),
                              gamma = c(0, 1.6),
                              num_trees = 1000)
```

### Fix overall sample size and tune % random sample and % honesty sample

#### Only have one effect modifier

```{r}
param_list = list(
  sample_fractions = seq(0.2, 0.5, by = 0.1),
  honesty_fractions = seq(0.3, 0.7, by = 0.1)
)

sim_result_sub1 <- run_simulation(sim_no = 500,
                                  param = param_list,
                                  sample_param = 600, 
                                  k = 1, d = 2,
                                  x_prob = c(0.5,0.7),
                                  beta = c(0, 0.1, 0.1),
                                  gamma = c(0, 1.6),
                                  num_trees = 1000)
```

#### Two effect modifiers X1, X2

```{r}
param_list = list(
  sample_fractions = seq(0.2, 0.5, by = 0.1),
  honesty_fractions = seq(0.3, 0.7, by = 0.1)
)

sim_result_sub2 <- tune_param(sim_no = 500, 
                             sample_size = 600, 
                             param = param_list, 
                             k = 2, d = 3,
                             x_prob = c(0.5,0.5,0.7),
                             beta = c(0, 0.1, 0.1,0.1),
                             gamma = c(0, 1.6),
                             num_trees = 1000)
```

### Show the results

```{r}
sim_result <- sim_result2
fig <- plot_ly(data = sim_result) %>%
  
  # Heterogeneity Captured
  add_trace(
    x = ~sample_size,
    y = ~heterogeneity_captured * 100,
    type = 'scatter',
    mode = 'lines+markers',
    name = 'Heterogeneity Captured',
    line = list(color = 'black', width = 2),
    marker = list(symbol = 'circle', size = 8, color = 'black'),
    hovertemplate = paste(
      'Sample Size: %{x}<br>',
      'Heterogeneity Captured: %{y:.2f}%<extra></extra>'
    )
  ) %>%
  
  # Success X
  add_trace(
    x = ~sample_size,
    y = ~success_ratio * 100,
    type = 'scatter',
    mode = 'lines+markers',
    name = 'Success X Tree',
    line = list(color = 'blue', width = 2),
    marker = list(symbol = 'circle', size = 8, color = 'blue'),
    hovertemplate = paste(
      'Sample Size: %{x}<br>',
      'Success X: %{y:.2f}%<extra></extra>'
    )
  ) %>%
  
  # Partial Success X
  add_trace(
    x = ~sample_size,
    y = ~partial_success_ratio * 100,
    type = 'scatter',
    mode = 'lines+markers',
    name = 'Partial Success X Tree',
    line = list(color = 'lightblue', width = 2),
    marker = list(symbol = 'circle', size = 8, color = 'lightblue'),
    hovertemplate = paste(
      'Sample Size: %{x}<br>',
      'Partial Success X: %{y:.2f}%<extra></extra>'
    )
  ) %>%
  
  # customize layout
  layout(
    title = '',
    xaxis = list(title = 'Sample Size', tickvals = sim_result$sample_size,
                 ticktext = ifelse(sim_result$sample_size == 0, "", sim_result$sample_size)),
    yaxis = list(title = 'Percentage (%)', range = c(0, 105)),
    hovermode = 'closest'
  )

fig
```

```{r}
# heterogeneity_captured
sim_result_sub <- sim_result_sub1
ggplot(sim_result_sub, aes(x = sample_fractions, 
               y = honesty_fractions, 
               fill = heterogeneity_captured)) +
  geom_tile(color = "white") +  
  scale_fill_viridis_c(option = "viridis", direction = 1) +
  labs(
    title = "Heatmap of Heterogeneity Captured",
    x = "Sample Fractions",
    y = "Honesty Fractions",
    fill = "Heterogeneity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)  
  )
```

```{r}
# success of pure X leaf
ggplot(sim_result_sub, aes(x = sample_fractions, 
               y = honesty_fractions, 
               fill = success_ratio)) +
  geom_tile(color = "white") +  
  scale_fill_viridis_c(option = "viridis", direction = 1) +
  labs(
    title = "Heatmap of Success Ratio of Pure Effect Modifier Leaf",
    x = "Sample Fractions",
    y = "Honesty Fractions",
    fill = "Success Ratio"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)  
  )
```
